---
title: "Getting started with landlock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with landlock}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# The output in this vignette depends on the kernel of the machine that
# builds it, so the code is shown with output captured on Linux 6.17
# (Landlock ABI 7, Ubuntu 24.04) instead of being run.
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

landlock confines R code to what it needs. You describe what a piece of code
may do as a *policy*; the package forks the R session, has the Linux kernel
enforce the policy on the child, evaluates the code there and hands back the
result. The session itself is never restricted.

## Evaluate code in a confined child

`eval_safe()` evaluates an expression in a forked child and returns its
value. Errors come back with their class; output is forwarded as it arrives;
`timeout` is a wall-clock limit that holds even inside C code.

```{r}
library(landlock)

set.seed(1)
eval_safe(mean(rnorm(1e6)))
#> [1] 4.690776e-05

tryCatch(eval_safe(stop("not here")), error = conditionMessage)
#> [1] "not here"

eval_safe(Sys.sleep(60), timeout = 2)
#> Error: timeout reached (2 sec)
```

Without a policy nothing is confined: this is `unix::eval_safe()`. A policy
adds the kernel layers:

```{r}
eval_safe(tryCatch(readLines("/etc/hostname"), warning = conditionMessage),
          policy = preset("numeric"))
#> [1] "cannot open file '/etc/hostname': Permission denied"
```

A denied file access surfaces in R as a warning ("Permission denied")
followed by an error ("cannot open the connection"). Warnings raised in the
child are not replayed in the session, as with `unix`, so catch them inside
the expression when you need the reason.

## Policies

A policy is a list you build with verbs:

```{r}
p <- policy() |>
  fs(read = c(R.home(), .libPaths(), "/usr"), tmp = TRUE) |>
  net() |>
  syscalls(deny = c(preset("dangerous"), preset("no_exec"))) |>
  caps() |>
  limits(memory = "2g", cpu = 60, nofile = 256)
p
#> <landlock policy> best effort 
#>   fs read   /opt/R/4.6.1/lib/R, /home/runner/work/_temp/Library, /opt/R/4.6.1/lib/R/library, /usr
#>   fs tmp    the call's own temporary directory (read and write)  
#>   tcp       bind: none; connect: none
#>   limits    as=2 GiB, cpu=60, nofile=256 
#>   syscalls  deny 63 calls, action errno 
#>   caps      keep none 
```

* `fs()` puts the filesystem under Landlock. Everything not listed is
  denied. `read`, `write` and `exec` are separate rights (`write` does not
  imply `read`); `rw` is both. A rule covers the path and everything below.
  `tmp = TRUE` adds the call's own scratch directory, the child's `TMPDIR`;
  the session's `tempdir()` is not granted, because the session may later
  trust what it finds there.
* `net()` restricts TCP `bind()` and `connect()` to the listed ports;
  `net()` alone blocks both.
* `scope()` keeps signals and abstract Unix sockets inside the sandbox.
* `syscalls()` denies system calls with a seccomp filter.
* `caps()` drops capabilities and sets `no_new_privs`.
* `limits()` sets resource ceilings; `ids()` switches user (root only);
  `apparmor()` changes the AppArmor profile.

Presets are policies to start from: `preset("numeric")` for code that only
computes, `preset("install")` for installing a source package,
`preset("plumber")` for serving HTTP. The system call sets
`preset("dangerous")`, `preset("no_exec")` and `preset("no_net")` are for
`syscalls()`.

## Reports, best effort and strict mode

Each layer depends on the kernel. When one is missing (an old kernel, a
container that blocks it, macOS) it is skipped, and the report says so. On a
kernel with Landlock ABI 3 (Linux 6.2 to 6.6):

```{r}
eval_safe(1, policy = preset("numeric"))
last_report()
#> <landlock report> Landlock ABI 3 
#>   landlock-fs     applied  ABI 3, 14 rules
#>   landlock-net    skipped  TCP rules need Landlock ABI 4; the kernel offers 3
#>   landlock-scope  skipped  scopes need Landlock ABI 6; the kernel offers 3
#>   seccomp         applied  deny 76 calls (errno EPERM); not on this architecture: socketcall; clone() with namespace flags refused
#>   caps            applied  bounding set kept (needs CAP_SETPCAP); effective, permitted, inheritable and ambient cleared
```

When a skipped layer is not acceptable, ask for strict mode and the
evaluation fails instead:

```{r}
eval_safe(1, policy = policy(best_effort = FALSE) |> net())
#> Error: Landlock: the kernel lacks a requested feature (ABI 3) and strict mode is on
```

`status()` tells you in advance what this machine offers.

## Running programs

`run()` does for a program what `eval_safe()` does for R code. Under a
filesystem policy the program needs `exec` permission on itself and on the
dynamic loader, which the kernel opens for execution too. The presets deny
`execve` outright, so build the policy for the program:

```{r}
p <- policy() |>
  fs(read = c("/usr", "/etc/ld.so.cache"), exec = c("/usr/bin", "/usr/lib"))
res <- run("cat", "/etc/hostname", policy = p)
res$status
#> [1] 1
rawToChar(res$stderr)
#> [1] "cat: /etc/hostname: Permission denied\n"

run("id", policy = preset("numeric"))
#> Error: cannot run 'id': Operation not permitted
```

## Confining the session itself

`confine()` applies a policy to the running session, irreversibly. Landlock
only restricts the calling thread and what it starts afterwards, so
`confine()` refuses when the session already runs other threads (a
multi-threaded BLAS, for example) unless you pass `force = TRUE`.
`restrict_self()`, `seccomp_deny()`, `caps_drop_all()` and `no_new_privs()`
are the thin, single-layer versions.

## Values from confined code

Under a policy the package treats the child as hostile: it never evaluates
what the child sends back. The value you get is still the child's data, and
R values can carry code (closures, environments with active bindings that
run when read). From code you do not trust, return plain data such as
vectors, lists and data frames.

## Moving from unix

landlock exports every function of `unix` 1.6.0 with the same arguments, in
the same order, with the same defaults. Replace `library(unix)` or `unix::`
with `library(landlock)` or `landlock::`; nothing changes until you pass a
`policy`. The differences:

* `eval_safe()` has a trailing `policy` argument, and records a report for
  `last_report()`.
* `unix` switches `tempdir()` and `interactive()` in the child by writing to
  R internals, which a package on CRAN may not do. Here `tmp` becomes the
  child's `TMPDIR` environment variable; `tempdir()` and `interactive()`
  keep the session's values, and standard input is `/dev/null`.
* `eval_safe(profile =)` changes the AppArmor profile through `/proc`, so it
  works without the 'RAppArmor' package or 'libapparmor'.
* `uid` without `gid` switches to the user's primary group and drops the
  caller's supplementary groups; `unix` keeps them.
* The default `tmp` is removed after the call.
* `status()`, `policy()`, `run()` and the confinement layers are new.
