---
title: "Worked examples"
---

```{r, include = FALSE}
# A pkgdown-only article: it is not part of the package tarball, and unlike
# the vignettes its code runs when the site is built. The site is built on a
# GitHub Actions Linux runner, so every output below is what landlock really
# did there.
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(landlock)
```

This article runs every example when the site is built. The machine that
built it:

```{r}
status()
```

On a kernel without Landlock, or on macOS, the same code runs; the layers the
kernel lacks are reported as skipped instead of enforced.

## Evaluate code you do not trust

A common case: an application receives R code, from a web form or an API,
and has to run it. `eval_safe()` runs it in a forked child, and a policy
decides what that child may do. The `"numeric"` preset allows computing and
nothing else.

```{r}
run_untrusted <- function(code, timeout = 5) {
  expr <- str2lang(code)
  eval_safe(eval(expr), policy = preset("numeric") |> limits(memory = "4g"),
            timeout = timeout)
}
```

Ordinary computation works:

```{r}
run_untrusted("summary(lm(mpg ~ wt + hp, data = mtcars))$r.squared")
```

Reading a file outside R's own files fails. Landlock refuses the open, which
R reports as a warning (the reason) and an error:

```{r, error = TRUE}
run_untrusted("readLines('/etc/passwd', n = 1)")
```

Warnings raised in the child stay in the child, as with `unix::eval_safe()`.
Catch them inside the expression to see the reason:

```{r}
run_untrusted("tryCatch(readLines('/etc/passwd', n = 1), warning = conditionMessage)")
```

Starting a program fails: the preset's seccomp filter refuses `execve`.

```{r}
run_untrusted("tryCatch(system('id', intern = TRUE), error = conditionMessage, warning = conditionMessage)")
```

So does opening a socket, which the filter refuses as well:

```{r}
run_untrusted("tryCatch(serverSocket(18080), error = conditionMessage)")
```

A loop that never ends is killed when the timeout passes, even when it is
stuck in C code:

```{r, error = TRUE}
run_untrusted("repeat {}", timeout = 2)
```

And a request for more memory than the limit fails inside the child, leaving
the session untouched:

```{r, error = TRUE}
run_untrusted("x <- numeric(1e9)")
```

What was enforced is in the report:

```{r}
last_report()
```

A layer the kernel cannot provide is reported as `skipped`. When that is not
acceptable, build the policy with `policy(best_effort = FALSE)` and the
evaluation fails instead of running with less protection.

## A policy for a task

Presets are a starting point. For a specific job, list exactly what it needs.
Here a function summarises a CSV file from an input directory and writes the
result to an output directory; it may read the input, write the output, and
nothing else of the user's.

```{r}
input <- file.path(tempdir(), "input")
output <- file.path(tempdir(), "output")
dir.create(input)
dir.create(output)
write.csv(mtcars, file.path(input, "cars.csv"), row.names = FALSE)

task_policy <- policy() |>
  fs(read = c(R.home(), .libPaths(), "/usr", input), write = output) |>
  net() |>
  syscalls(deny = c(preset("dangerous"), preset("no_exec"), preset("no_net"))) |>
  caps() |>
  limits(memory = "4g", cpu = 30)
task_policy
```

```{r}
summarise_file <- function(name) {
  eval_safe({
    d <- read.csv(file.path(input, name))
    s <- aggregate(mpg ~ cyl, data = d, FUN = mean)
    write.csv(s, file.path(output, "summary.csv"), row.names = FALSE)
    nrow(s)
  }, policy = task_policy)
}

summarise_file("cars.csv")
read.csv(file.path(output, "summary.csv"))
```

The same code cannot reach anything else. Writing into the input directory,
or reading the output it just wrote, is refused: `write` and `read` are
separate rights.

```{r}
eval_safe(file.create(file.path(input, "planted.csv")), policy = task_policy)
eval_safe(file.exists(file.path(output, "summary.csv")), policy = task_policy)
eval_safe(tryCatch(readLines(file.path(output, "summary.csv")), warning = conditionMessage),
          policy = task_policy)
```

`file.exists()` still answers: Landlock governs opening files, not looking at
their metadata.

## Running a program

`run()` applies a policy to a program instead of R code. The program needs
permission to execute itself and the dynamic loader (the kernel opens both
for execution), and to read its libraries.

```{r}
words <- file.path(input, "words.txt")
writeLines(c("pear", "apple", "fig"), words)

sort_policy <- policy() |>
  fs(read = c("/usr", "/etc/ld.so.cache", words), exec = c("/usr/bin", "/usr/lib")) |>
  net()

res <- run("sort", words, policy = sort_policy)
res$status
cat(rawToChar(res$stdout))
```

The same program on a file the policy does not list:

```{r}
res <- run("sort", "/etc/passwd", policy = sort_policy)
res$status
cat(rawToChar(res$stderr))
```

## Coming from unix

Code written for the 'unix' package runs unchanged:

```{r}
eval_safe(rlimit_cpu()$cur, rlimits = c(cpu = 10))
eval_fork(getppid() == getpid())
```

and gains confinement by adding `policy`:

```{r}
eval_safe(sum(1:10), rlimits = c(cpu = 10), policy = preset("numeric"))
```

## Values from confined code

Under a policy the package never evaluates what the child sends back, but
the value you receive is the child's data. R values can carry code:
closures, and environments whose active bindings run when read. From code
you do not trust, return plain data such as vectors, lists and data frames,
as `run_untrusted()` above does.
